A family of accessibility measures

### 2024-11-12 paper/manuscript.qmd
draft of accessibility formulation

## Entropy maximising: the gravity-model as statistical averages

Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling]. Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].

@@

### Unconstrained interaction case

@@

### Total constrained interaction case

@@

### Singly-constrained

@@

### Doubly-constrained case

@@

### Mean absolute percentage error (MAPE) between all cases

@@

#total constrained
((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

@@

@@

$$
V_{ij} = k O_i D_j f(c_{ij})
$${#eq-access}

Where $V_{ij}$ represents a $n$ x $m$ matrix of potential reachable opportunities between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In other words, $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data. Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.

Like with @eq-phys-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel). To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used. These pieces of information can constrain $V_{ij}$ as follows, ensuring it reflects this macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.

$$
\sum_j V_{ij} = O_i
$${#eq-constraint1-access}
\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the opportunities originating $O_i$ at each $i$ in the region.

$$
\sum_i V_{ij} = D_j
$${#eq-constraint2-access}
\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the opportunities ending $D_i$ at each $j$ in the region.

$$
\sum_{i}\sum_{j} V_{ij} c_{ij} = C
$${#eq-constraint3-access}
\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.

For all FIVE cases, @eq-constraint3-access holds <!--in the event C is known, it could be used to identify parameters in f(c_ij)... though model calibration is probably a better idea. --> but are differentiated like:
- neither @eq-constraint1-access or @eq-constraint2-access hold -- the unconstrained case;
- @eq-constraint1-access holds -- the production constrained case (origin-constrained, a singly-constrained case); 
- @eq-constraint2-access holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
- @eq-constraint1-access and @eq-constraint2-access both hold simultaneously -- the production-attraction constrained case (doubly-constrained).
- @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->

Now how does this translate to accessibility?

Opportunities (say, grocery stores):
```{r}
D_j <- c(7, 3, 2)
```

## Unconstrained accessibility 
(no proportionality constant):
$$
S_i = \sum_j\frac{D_j}{C_{ij}}
$$

```{r}
S_i <- (1/C) %*% D_j
```

These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
```{r}
sum(S_i)
```

##Total flow constrained accessibility, needs a balancing factor:
```{r}
k <- sum(D_j)/sum(S_i)

k
```

## Unconstrained accessibility (no proportionality constant):
$$
S_i = k\sum_j\frac{D_j}{C_{ij}}
$$

```{r}
S_tot <- k * S_i

S_tot
```

Now check the sum:
```{r}
sum(S_tot)
```

Neither @eq-constraint1-access or @eq-constraint2-access hold, so $V_ij'$ has no opportunity constraints!

$$
V_{ij}' = k O_i D_j f(c_{ij})
$${#eq-access-unconstrained}

Where $k$ is two sets of balancing factors, $A_i'$ and $B_j'$, and both are equal to 1.

## Singly-constrained case (origin-constrained)

Spatial availability; origin-constrained.

Origins in human spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998; for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024

$$
V_{ij}'' = k O_i D_j f(c_{ij})
$${#eq-access-singlyconstrained}

<!--HOW DO WE FIT OUR EXISTING CONSTRAINTS INTO THESE BALANCING FACTORS..? $k O_i f(c_{ij}) = the balancing factors..?-->
$$
V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
$$
Population balancing factor:
$$
F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}
$$
Travel cost balancing factor:
$$
F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}
$$
## Singly-constrained case (destination-constrained)

Origins are in retail models, decision model for shops to optimally spatially locate to max. their potential consumer interaction. Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965. Spatial availability as another example.

@@

## Hybrid singly-constrained case: 
<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit??-->
Balanced floating catchment areas (not truly doubly constrained)

@@

>>>>

**Entropy maximising: the gravity-model as statistical averages:**
Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling]. Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].

@@

## Unconstrained interaction case

@@

## Total constrained interaction case

@@

## Singly-constrained

@@

## Doubly-constrained case

@@

## Mean absolute percentage error (MAPE) between all cases

@@

#unconstrained -- rescaled!
W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums()
W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums()

W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums()
W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums()

((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01)
```

```{r}
#total constrained
((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

@@

- doubly c is lowest error
- singly constrained error is mid
- total constrained has high error
- notably, unconstrained unscaled is obviously much higher in error - there are no constraints! But if you scale the matrices, the error is both low (interesting!! why is this?). However as mentioned, the unconstrained case is hard to interpret literally -- in what situations would k=1..?

@@

$$
V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
$${#eq-access}

Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.

Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.

$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}

$$
\sum_i\sum_j V_{ij} c_{ij} = C
$${#eq-constraint0_5-access}
\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.

$$
\sum_j V_{ij} = W_i^{(1)}
$${#eq-constraint1-access}
\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

$$
\sum_i V_{ij} = W_j^{(2)}
$${#eq-constraint2-access}
\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

All four <!--five?--> cases are differentiated like:
- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
- The total constrained case: only @eq-constraint0-gravitymodel hold
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
- HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->

We can use the same trip matrix, but reformulate its meaning. Let's say $W_i^{(1)}$ is population at $i$ who seek opportunities, lets recast this variable as $P_i$. Then let's say $W_j^{(2)}$ are the opportunities being sought, recast to $O_j$. For this simple example, we can say that the opportunities are the population's jobs. In our region, each person has a job -- but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost ($C$ and associated travel propensity $1/C$). (let's pretend our 'known_trip_matrix' is unknown).

```{r}
known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
W_i <- known_trip_matrix |> rowSums()
W_j <- known_trip_matrix |> colSums()
T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
knowns_longer <- cbind(known_trip_matrix,W_i)
knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))

## Unconstrained accessibility 
(no proportionality constant, k=1, no constraints hold - hard to interpret):
$$
S_{ij} = P_i O_j f(c_{ij})
$$
In practice, $S_i$ is calculated for each $i$ so:
$$
S_{i} = P_i \sum_j O_j f(c_{ij})
$$
These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).

Looks like Hansen, but multiplied by P_i.

## Total flow constrained accessibility

This needs a balancing factor, $k$ is replaced with $K$:
$$
S_{ij}' = K P_i O_j f(c_{ij})\\
S_{i}' = K P_i \sum_j O_j f(c_{ij})
$$

This approach is not seen in accessibility literature...

## Singly-constrained case

Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$. For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
$$
V_{ij}''^{(1)} = A_i P_i O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_j P_i O_j f(c_{ij})\\

V_{i}''^{(1)} = A_i P_i \sum_j O_j f(c_{ij})\\
V_{i}''^{(2)} = P_i \sum_j B_j O_j f(c_{ij})\\

$${#eq-access-singlyconstrained-origin}

\noindent Where $A_i$ and $B_j$ are:
$$
A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
$$
For the origin-constrained, $V_{i}''^{(1)}$ are the number of opportunities (like jobs) that can be reached at $i$, and $V_{i}''^{(2)}$ are the number of opportunities (like people) that can be reached at $i$

In terms of origin-constrained -- spatial availability is explicitly so [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024]. Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general [].

Regarding destination-constrained, has it's origins are in retail models. e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction. Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.

For instance, the most recent example spatial availability is formulated like so:
$$
V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
$$
Where the $F^p_{ij}$ Population balancing factor is:
$$
F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}
$$
And the $F^c_{ij}$ travel cost balancing factor:
$$
F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}
$$
It can be reformulated to fit into the singly-constrained (origin case) gravity model as the denominators in both those balancing factors cancel out:
$$
V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}\\

V_{ij}''^{(1)} =O_j \frac{1}{\sum_j O_j f(c_{ij})} P_i  f(c_{ij})\\
$$

@@

## Hybrid singly-constrained case: 
<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here-->
Balanced floating catchment areas (not truly doubly constrained)

@@

<!--TO DISCUSS:
-- should we use entropy maximisation to actually derive the models? I'm hazy on that.
  -- relatedly, in the unconstrained case... is k=1 or does k just not exist..? I assume k=1 and say that the interpretation in reality is confusing. a consistent derivation of all cases can help address this.
  -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->
